Title
========================================================

# Synopsis

# Data Processing
```{r cache=TRUE}
storm <- read.csv("repdata-data-StormData.csv.bz2")
numObs <- dim(storm)
```

According to the provided documentation, the possible events should be 48. Unfortunately, perhaps because of a not enough guided data entry procedure, across all `r numObs` observations the number of possible events is much higher, in fact it is almost 1000:
```{r}
length(levels(storm$EVTYPE))
```
Anyway, the questions that we are asked to answer are the following:

* Across the United States, which types of events (as indicated in the EVTYPE variable) are most harmful with respect to population health?
* Across the United States, which types of events have the greatest economic consequences?

So, even if our original data contain more than 900000 samples, we can restrict our analysis to those samples that have at least one of the following columns not equal to 0:
* FATALITIES
* INJURIES
* PROPDMG
* CROPDMG

```{r}
storm <- storm[(storm$FATALITIES!=0) | (storm$INJURIES!=0) | (storm$PROPDMG !=0) | (storm$CROPDMG !=0),]
```

Furthermore, only a subset of the available columns are necessary for our analysis, so we can subset our original dataframe like this:

```{r}
storm <- data.frame(storm$EVTYPE, storm$FATALITIES, storm$INJURIES, storm$PROPDMG, 
    storm$PROPDMGEXP, storm$CROPDMG, storm$CROPDMGEXP, stringsAsFactors = FALSE)
str(storm)
```

Now we need to normalize the *PROPDMGEXP* and *CROPDMGEXP* because, in the form they are given, they are useless for us.

```{r cache=TRUE}
storm$storm.PROPDMGEXP<-as.character(storm$storm.PROPDMGEXP)
storm$storm.CROPDMGEXP<-as.character(storm$storm.CROPDMGEXP)

storm$storm.PROPDMGEXP[(storm$storm.PROPDMGEXP == "")] <- 0
storm$storm.PROPDMGEXP[(storm$storm.PROPDMGEXP == "+") | (storm$storm.PROPDMGEXP == 
    "-") | (storm$storm.PROPDMGEXP == "?")] <- 1
storm$storm.PROPDMGEXP[(storm$storm.PROPDMGEXP == "h") | (storm$storm.PROPDMGEXP == 
    "H")] <- 2
storm$storm.PROPDMGEXP[(storm$storm.PROPDMGEXP == "k") | (storm$storm.PROPDMGEXP == 
    "K")] <- 3
storm$storm.PROPDMGEXP[(storm$storm.PROPDMGEXP == "m") | (storm$storm.PROPDMGEXP == 
    "M")] <- 6
storm$storm.PROPDMGEXP[(storm$storm.PROPDMGEXP == "B")] <- 9

storm$storm.CROPDMGEXP[(storm$storm.CROPDMGEXP == "")] <- 0
storm$storm.CROPDMGEXP[(storm$storm.CROPDMGEXP == "+") | (storm$storm.CROPDMGEXP == 
    "-") | (storm$storm.CROPDMGEXP == "?")] <- 1
storm$storm.CROPDMGEXP[(storm$storm.CROPDMGEXP == "h") | (storm$storm.CROPDMGEXP == 
    "H")] <- 2
storm$storm.CROPDMGEXP[(storm$storm.CROPDMGEXP == "k") | (storm$storm.CROPDMGEXP == 
    "K")] <- 3
storm$storm.CROPDMGEXP[(storm$storm.CROPDMGEXP == "m") | (storm$storm.CROPDMGEXP == 
    "M")] <- 6
storm$storm.CROPDMGEXP[(storm$storm.CROPDMGEXP == "B")] <- 9

storm$storm.PROPDMGEXP <- as.integer(storm$storm.PROPDMGEXP)
storm$storm.CROPDMGEXP <- as.integer(storm$storm.CROPDMGEXP)
```

Having normalized the exponents necessary to correctly interpretate the quantities given, we can now calculate the total damage in a new column:

```{r cache=TRUE}
storm$total.damage <- storm$storm.PROPDMG * 10^storm$storm.PROPDMGEXP + storm$storm.CROPDMG * 10^storm$storm.CROPDMGEXP
```

At this point we use the ddply function of the *plyr* package to calculate the sum of fatalities, injuries and total damages for each of the event types in the subsetted dataframe:
```{r cache=TRUE}
require(plyr)
stormSummary <- ddply(storm, .(storm.EVTYPE), summarize, fatalities=sum(storm.FATALITIES), injuries=sum(storm.INJURIES), damages=sum(total.damage))
```



# Results
Having prepared the dataframe with the summarized data that we need, all we have to do is extract the required statistics.
Also, we will use the *ggplot2* library plotting system.

```{r}
require(ggplot2)
```

### Injuries

```{r cache=TRUE}
stormInjuries <- head(stormSummary[order(stormSummary$injuries, decreasing=TRUE),])[,c(1,3)]
print(stormInjuries)
ggplot(data=stormInjuries,aes(storm.EVTYPE, injuries)) + geom_bar(stat="identity") + theme(axis.text.x=element_text(angle=90, hjust=1)) + xlab("Event type") + ylab("Injuries") + ggtitle("Top Injuries Counts")
```

### Fatalities

```{r cache=TRUE}
stormFatalities <- head(stormSummary[order(stormSummary$fatalities, decreasing=TRUE),])[,c(1,2)]
print(stormFatalities)
ggplot(data=stormFatalities,aes(storm.EVTYPE, fatalities)) + geom_bar(stat="identity") + theme(axis.text.x=element_text(angle=90, hjust=1)) + xlab("Event type") + ylab("Fatalities") + ggtitle("Top Fatalities Counts")
```

### Damages
```{r cache=TRUE}
stormDamages <- head(stormSummary[order(stormSummary$damages, decreasing=TRUE),])[,c(1,4)]
print(stormDamages)
ggplot(data=stormDamages,aes(storm.EVTYPE, damages)) + geom_bar(stat="identity") + theme(axis.text.x=element_text(angle=90, hjust=1)) + xlab("Event type") + ylab("Damages") + ggtitle("Top Damages Counts")
```